# Introduction and Context

```{r}
#| label: load-pkg

library(tidyverse)
library(tidymodels)
library(knitr)
library(reshape)
library(ggplot2)
library(dplyr)
library(lme4)
library(MASS)
library(car)
library(MuMIn)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(broom.mixed)
library(forcats)
library(sjPlot)

theme_set(theme_minimal(base_size = 11))
```

```{r}
#| label: load-data
#| message: false

gardnerjanson <- read_csv(here::here("processed-data/gardnerjanson.csv"))

gardnerjanson_museums <- read_csv(here::here("processed-data/gardnerjanson_museums.csv"))
```

```{r}
#| label: lmm_prep

gardnerjanson_museums_mod <- gardnerjanson_museums %>%
  filter(!startsWith(artist_name, "N/A")) %>%
  mutate(artist_race_nwi = if_else(artist_race == "White", "White", "Non-White"))

gardnerjanson_museums_mod <- gardnerjanson_museums_mod %>%
  mutate(artist_nationality_other = factor(artist_nationality_other,
    levels = c("American", "French", "Other", "British", "German", "Spanish")
  ))
```

```{r}
#| label: lmm_full

lmm_full <- lmer(log(space_ratio_per_page_total) ~ artist_race_nwi
                       + artist_ethnicity 
                       + artist_gender 
                       + artist_nationality_other
                       + moma_count_to_year
                       + whitney_count_to_year +
                         (1 | artist_name),
            data = gardnerjanson_museums_mod,
    REML = FALSE)
```

```{r}
#| label: lmm_test

lmm <- lmer(log(space_ratio_per_page_total) ~ artist_race_nwi
                       + artist_ethnicity 
                       + artist_gender 
                       + artist_nationality_other
                       + moma_count_to_year
                       + whitney_count_to_year
                       + artist_nationality_other*moma_count_to_year
                       + artist_race_nwi*moma_count_to_year
                       + artist_ethnicity*moma_count_to_year
                       + artist_race_nwi*whitney_count_to_year
                       + artist_ethnicity*whitney_count_to_year
                  + (1 | artist_name),
            data = gardnerjanson_museums_mod,
    REML = FALSE)
```

```{r}
#| label: lmm_step

final_model <- lmerTest::step(lmm)
```

```{r}
#| label: lmm

lmm <- lmer(log(space_ratio_per_page_total) ~ artist_nationality_other
                       + moma_count_to_year
                       + artist_nationality_other*moma_count_to_year
                        + (1 | artist_name),
            data = gardnerjanson_museums_mod,
    REML = FALSE)
```

## Inspiration for Research

Heading into the second semester of my junior year, I scheduled a zoom meeting with my now undergraduate thesis advisor, Prof. Hans van Miegroet, to ask a few questions about my aspirations after college.
He had been my professor for his course, "History of Art Markets," a learning experience at Duke that has permanently altered my worldview.
The meeting began with me expressing my enjoyment of the course, and the appreciation of the discussion of transparency in the buying and selling of art.
I asked his opinion of whether I should stick to studying architectural history with hopes of going to architecture school, or pivot to do further research under him in his graduate course, "Arts and Markets" that I had enrolled in for that Spring.
Towards the end of the meeting, he asked me what I found most interesting about art, to which I replied, "Ever since I took my first Art History course, I've always been curious as to why I am introduced to the works I am. Who is choosing what I study versus what I don't study?" I explained to him.
I had initially learned the story of art through Marilyn Stokstad's, *Art History* my final year of high school.
At the end of that school year, I inquired to my teacher, Carolyn Paczkowska, in a class discussion, "Why are we studying the works that we have?" The class discussed gatekeepers of art and information and later ended in the conclusion of uncertainty.
Prof. van Miegroet looked at me through the computer screen and stated I had found my research question for my Undergraduate Honors Thesis, that I had found my why.

My research began solely looking at how art history textbooks changed through time, then developed into looking at using a linear mixed-effect model with a log transformation on my outcome variable to infer which external variables would if at all work to predict the magnitude of space (of text and of the figure of their work or works) given to an artist in a given book.

## Why *Gardner's Art Through the Ages*

First published in 1926 by Helen Gardner, *Gardner's History of Art* is widely considered as the first single-volume survey of the world's art.[^intro-1]
Her introductory survey text far surpassed other available works in readability, breadth of coverage, and wealth of illustration.
[^intro-2] The latest edition, the sixteen, of the art history survey text, was published in 2020
. *Gardner's Art Through the Ages* has the longest publishing history and evolution that displays the changing nature of an art history survey text
.

[^intro-1]: Kader, Themina.
    "The Bible of Art History: Gardner's Art through the Ages." Studies in Art Education 41, no. 2 (Winter, 2000): 176.

[^intro-2]: britannica but look if info is on lee's art historian dictionary

Gardner's initial success can be quantified by the statistic that the first two editions, published a decade apart, as stated by art education researcher from Pennsylvania State University, Kader Themina, "sold 260,000 copies, a testimony to the popularity of her book." Additionally, as highlighted by a team of researchers at Purdue in as recent as 2020, "Art history instructors use and rely on Gardner's global coverage..."[^intro-3] In a footnote in that same publication, an unnamed art historian, in an email to the authors, states "Gardner's Art Through the Ages is, by a wide margin, the best-selling introduction to the history of art in the English language and the number one test in North American high school and university courses."

[^intro-3]: purdue 14

Add any more sales information if you can, and state if you cannot find anything else on it.

Though it is unclear where this information is coming from, it is clear that *Gardner's Art Through the Ages* is a phenomenal proxy of how the overall narrative of the history of art is changing through time.

### Discourse of Authorship

**(**Discourse of dominance through time of the text, find and add sources and if available sales information through time)

**TO DO: REACH OUT FOR SALES DATA TO PUBLISHERS**

Britannica

-   The lack of a [comprehensive](https://www.merriam-webster.com/dictionary/comprehensive) single-volume textbook on art history prompted Gardner to write one herself, and the resulting Art Through the Ages (1926) far surpassed other available works in readability, breadth of coverage, and wealth of illustration. It remained a widely used text for decades. In 1932 she published Understanding the Arts, aimed at a wide general audience. A second edition of Art Through the Ages, greatly expanded, appeared in 1936; **the first two editions sold more than 260,000 copies.** Gardner had been named an assistant professor in the Art Institute school in 1929, and in 1933 she became a professor and head of the department of art history. She retired from the Art Institute school in 1943. Despite declining health she managed to complete work on the manuscript of a third edition of Art Through the Ages (published in 1948).

Sources Found to Looking Into:

-   Franciscono, Marcel.
    1977. "History, Textbooks, and Art: Reflections on a Half Century of Helen Gardner's "Art through the Ages"." Critical Inquiry 4 (2) (Winter): 285. https://login.proxy.lib.duke.edu/login?u rl=https://www.proquest.com/scholarly-journals/history-textbooks-art-reflections-on-half-century/docview/1297337757/se-2?a ccountid=10598.
-   Jaffee, Barbara. "9. "Gardner" Variety Formalism: Helen Gardner and Art through the Ages" In Partisan Canons edited by Anna Brzyski, 203-224. New York, USA: Duke University Press, 2007. https://doi.org/10.1515/9780822390374-010.

+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| Edition  | Year     | Author(s)                                                                                       | Title                                                                 | Publisher                                               |
+==========+==========+=================================================================================================+=======================================================================+=========================================================+
| 1        | 1926     | Helen Gardner                                                                                   | Art Through the Ages; An Introduction to Its History and Significance | New York, Brace, Harcourt                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 2        | 1936     | Helen Gardner                                                                                   | Art Through the Ages; An Introduction to Its History and Significance | New York, Brace, Harcourt                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 3        | 1948     | Helen Gardner                                                                                   | Art Through the Ages                                                  | New York, Brace, Harcourt                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 4        | 1959     | Helen Gardner; revised by Sumner M. Crosby and the Dept. of the History of Art, Yale University | Art Through the Ages                                                  | New York, Harcourt, Brace                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 5        | 1970     | Revised by Horst de la Croix, Richard G. Tansey                                                 | Gardner's Art Through the Ages                                        | New York, Harcourt, Brace                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 6        | 1975     | Helen Gardner; revised by Horst de la Croix, Richard G. Tansey                                  | Gardner's Art Through the Ages                                        | New York: Harcourt Brace Jovanovich                     |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 7        | 1980     | Helen Gardner; revised by Horst de la Croix, Richard G. Tansey                                  | Gardner's Art Through the Ages                                        | Harcourt Brace Jovanovich, New York                     |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 8        | 1986     | Horst de la Croix, Richard G. Tansey                                                            | Gardner's Art Through the Ages                                        | Harcourt Brace Jovanovich, San Diego, CA                |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 9        | 1991     | Horst de la Croix, Richard G. Tansey, Diane Kirkpatrick                                         | Gardner's Art Through the Ages                                        | San Diego: Harcourt Brace Jovanovich                    |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 10       | 1996     | Richard G. Tansey, Fred S. Kleiner                                                              | Gardner's Art Through the Ages                                        | Fort Worth, TX: Harcourt Brace                          |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 11       | 2001     | Fred S. Kleiner, Christin J. Mamiya, Richard G. Tansey                                          | Gardner's Art Through the Ages                                        | Fort Worth TX: Harcourt College Publishers              |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 12       | 2005     | Fred S. Kleiner, Christin J. Mamiya                                                             | Gardner's Art Through the Ages                                        | Thomson/Wadsworth, Belmont, CA                          |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 13       | 2009     | Fred S. Kleiner                                                                                 | Gardner's Art Through the Ages: A Global History                      | Boston, Thomson/Wadsworth                               |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 14       | 2013     | Fred S. Kleiner                                                                                 | Gardner's Art Through the Ages: A Global History                      | Australia ; United States : Wadsworth, Cengage Learning |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 15       | 2016     | Fred S. Kleiner                                                                                 | Gardner's Art Through the Ages: A Global History                      | Boston, MA : Cengage Learning                           |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+
| 16       | 2020     | Fred S. Kleiner                                                                                 | Gardner's Art Through the Ages: A Global History                      | Boston, MA, US : Cengage Learning                       |
+----------+----------+-------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+---------------------------------------------------------+

: Edition Number, Year of Publication (as listed per Edition), Title, Authorship, and Publisher Over Time of All Cataloged Editions of *Gardner's Art Through the Ages*. {#tbl-gardner-editions tbl-colwidths="\[8,7,30,30,25\]"}

+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Author                             | Edition (Year)                                                              | Description                                                                                                                                                                                                                                                                              |
+====================================+=============================================================================+==========================================================================================================================================================================================================================================================================================+
| Sumner McKnight Crosby (1909-1982) | 4 (1959)                                                                    | Sumner McKnight Crosby was an American Medievalist architectural historian. He was the principal scholar of St-Denis and chair of the Department of Art History, Yale University, 1947-1953.[^intro-4]                                                                                   |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Horst de la Croix                  | 5 (1970), 6 (1975), 7 (1980), 8 (1986), 9 (1991)                            | An Art Historian and professor at San Jose State University from 1957 until his retirement in the early 1980s.[^intro-5] He was born in Berlin and came to the U.S. in 1935. He obtained his B.A., M.A., and Ph.D. in Art History from the University of California, Berkeley.[^intro-6] |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Richard G. Tansey (1919-1998)      | 5 (1970), 6 (1975), 7 (1980), 8 (1986), 9 (1991), 10 (1995)                 | An Art Historian and professor at San Jose State University from 1947-1961. He then taught at UCLA for a year, only to return to San Jose State University from 1962-1980. He received his B.A., M.A., and Ph.D. from Harvard.[^intro-7]                                                 |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Diane Kirkpatrick (1933 - present) | 9 (1991)                                                                    | She received her B.A. from Vassar College in 1955, her MA and Ph.D. in Art History from the University of Michigan in 1965 and 1969. She joined the faculty at the University of Michigan in 1968 and retired in May of 2000.[^intro-8]                                                  |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Fred S. Kleiner                    | 10 (1995), 11 (2001), 12 (2005), 13 (2009), 14 (2013), 15 (2016), 16 (2020) | He received his B.A. from the University of Pennsylvania, then his M.A. and Ph.D. in Art History from Columbia University.[^intro-9] He was a professor of the History of Art and Architecture and Archaeology at Boston University from 1978 - the present.[^intro-10]                  |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Christin J. Mamiya                 | 11 (2001), 12 (2005)                                                        | She received her B.A. from Yale University, and her M.A. and Ph.D degrees in Art History from UCLA. She was a professor at the University of Nebraska-Lincoln from 1987-2019.[^intro-11]                                                                                                 |
+------------------------------------+-----------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: Authors of *Gardner's Art Through the Ages* {#tbl-gardner-authors tbl-colwidths="\[25, 13, 62\]"}

[^intro-4]: https://arthistorians.info/crosbys

[^intro-5]: Little to no information available about Horst de la Croix

[^intro-6]: Staff Report, Mercury News.
    "H. DE LA CROIX, SJS PROFESSOR OF ART HISTORY." \*San Jose Mercury News (CA)\*, March 5, 1992: 3B.
    \*NewsBank: America's News -- Historical and Current\*.
    https://infoweb.newsbank.com/apps/news/document-view?p
    =AMNEWS&docref=news/0EB7195C84C7A20D.

[^intro-7]: https://scholarworks.sjsu.edu/erfa_bios/240/

[^intro-8]: http://faculty-history.dc.umich.edu/faculty/diane-m-kirkpatrick/memoir

[^intro-9]: https://www.bu.edu/archaeology/profile/kleiner/

[^intro-10]: https://prabook.com/web/fred.kleiner/3746660

[^intro-11]: <https://arts.unl.edu/art/faculty/christin-mamiya>

### Literature Review

"Gardner" Variety Formalism: Helen Gardner and *Art through the Ages*

Barbara Jaffee

https://editions.lib.umn.edu/panorama/wp-content/uploads/sites/14/2016/11/Jaffee-Art-Historys-Other-Global-Moment-Chicago-1948.pdf

## Why *Janson's History of Art*

Such a text is claimed to be the most influential art history survey through time by myriad art historians such as Jeffery Wiedman, Zoë Ingalls, John Russell, Alexandra Peers, Elizabeth Sears and Charlotte Schoell-Glass to name a few.
Alexandra Peers in her publication for ARTnews.com in February of 2006, claims that "it was Janson who, more than any other art historian, pioneers the 'in and out' celebrity model of art history. There were artists who matter, he argued, and those who didn't."[^intro-12] Janson's formation of his history of art is not only arguably the most dominant art history survey over time, but also he holds a reputation of being a gatekeeper of art history.
He states at the end of his introduction in the first and second editions that after having read his text, one "shall have joined the active minority that participates directly in shaping the course of art in our time."[^intro-13] He recognizes his role as an individual who shapes the narrative of the history of art, while convincing the reader that after having read his text, one has the agency to be a gatekeeper alongside him
. Such agency is only granted once one understands his digestion of the most important works and artists through time
.

[^intro-12]: Alexandra Peers, "Canon Fodder," Artnews.com, (February 1, 2006), https://www.artnews.com/art-news/news/canon-fodder-135/.

[^intro-13]: H. W. Janson and Dora Jane Janson, (1963),  History of Art: A Survey of the Major Visual Arts from the Dawn of History to the Present Day, First Edition, Second Printing, New York: Harry N. Abrams: 17.

It would be most comprehensive if we were to have access to the total number of sales per edition of Janson's History of Art, to use as tangible data to show the relevance and importance of the text through time.
Both publishers, Harry N. Abrams and Prentice-Hall (now Pearson Prentice-Hall) have declined sharing sales information in the aggregate.
By leafing through various scholarly publications, there is imperfect data of sale information worth noting when discussing the dominance of Janson's History of Art through time.
The text was at its height of sales between when it was initially released in 1962 through H.
W.
"Peter" Janson's death in 1982 (refer to @tbl-gardner-editions for complete information of each edition).
Art historian, Patricia Hills cites in her publication in Artforum, "in 1973 the Janson text had 46% of the market, while Gardner's 5th edition had 24%, followed by Gombrich's, 8.5%, Cleaver with 3% and miscellaneous 'other' with 18.5%."[^intro-14] At the time such a report had been made, the first edition (1962), the first edition revised and enlarged (1969) had been released.
Additionally, it was written by John Russell of the New York Times in October of 1982 that "well over two million copies have been sold."[^intro-15] Russell does not specify whether these sales numbers are solely the first edition of Janson's History of Art, or the first edition as well as the first edition revised and enlarged and the second edition (1977)
. He also added that the text had been "translated in 14 languages and is widely regarded as both an essential teaching instrument and a book that can be read and looked at with continuous enjoyment."[^intro-16] Here, Russell perpetuates the significance of Janson's survey, as well as providing information about the breadth of Janson's text, reaching more than solely English-reading audience
s.

[^intro-14]: Market Research Report on Introductory Art History, dated March 15, 1972, prepared for McGraw-Hill College Textbook Division and Follow-up Report, dated Spring, 1973.
    Citation from Patricia Hills, "Art History Textbooks: The Hidden Persuaders," Artforum, (Summer 1976), https://www.artforum.com/print/197606/art-history-textbooks-the-hidden-persuaders-69128.

[^intro-15]: John Russell, "Prof H. W Janson is Dead at 68; Wrote Best-Selling 'History of Art,'" The New York Times, (October 3, 1982), https://www.nytimes.com/1982/10/03/nyregion/prof-h-w-Janson-is-dead-at-68 -wrote-best-selling-history-of-art.html.

[^intro-16]: John Russell, "Prof H. W Janson is Dead at 68; Wrote Best-Selling 'History of Art,'" The New York Times, (October 3, 1982), https://www.nytimes.com/1982/10/03/nyregion/prof-h-w-Janson-is-dead-at-68 -wrote-best-selling-history-of-art.html.

Conversely, Zoë Ingalls in a publication in August of 1995 discussing Janson's relevance through time, cites that "the first edition sold more than a million copies."[^intro-17] Though there are informational discrepancies, it is clear that Janson's History of Art was prolific as a survey text of art for over 20 years.
The narrative remained significant from the third (1986) through the fifth edition (1995), which were all revised by Anthony "Tony" F.
Janson, son of Peter and Dora Jane.
Ingalls states that "the fourth edition, published in 1991, sold 21,000 copies the first year and an average of 11,000 copies a year in the past four years. The new fifth edition, has sold more than 13,000 copies since its March release."[^intro-18] She continues, "Although its sales have been eclipsed over the last 10 years by another perennial favorite, R.H. Gardner's Art Through the Ages, Janson is still the standard in many people's minds."[^intro-19] Notably, she does not mention any total sales information about Gardner's survey text, just that over the past decade (1985-1995), its sales trumped that of Janson'
s.

[^intro-17]: Zoë Ingalls, "A Son Revises His Father's Classic Art-History Textbook," The Chronicle of Higher Education, (August 11, 1995), https://www.chronicle.com/article/a-son-revises-his-fathers-classic- art-history-textbook/.

[^intro-18]: Zoë Ingalls, "A Son Revises His Father's Classic Art-History Textbook," The Chronicle of Higher Education, (August 11, 1995), https://www.chronicle.com/article/a-son-revises-his-fathers-classic- art-history-textbook/.

[^intro-19]: Zoë Ingalls, "A Son Revises His Father's Classic Art-History Textbook," The Chronicle of Higher Education, (August 11, 1995), https://www.chronicle.com/article/a-son-revises-his-fathers-classic- art-history-textbook/.

After the sixth edition (2001) and the sixth edition revised (2004) the last with authorship by Tony Janson, Harry N. Abrams and Prentice Hall were unsatisfied by the total sales, as well as the overall structure and content of the text as they were criticized for having lost touch with young readers.[^intro-20]
At the same moment, Harry N. Abrams sold all rights to the newly formed Pearson Prentice Hall, who recruited new authorship of six various art historians across the United States: Penelope J. E. Davies, Walter B. Denny, Frima Fox Hofrichter, Joseph Jacobs, Ann M. Roberts, and David L. Simon (refer to Table @tbl-janson-editions).
Such change was received negatively by art historians, which is reflected in the text's further decline in sales.
[^intro-21] It has been estimated by Bookauthority.com (a cite whose source for this information is very unclear) that the eighth edition revised, published in 2015 has had 2,000 total copies sold, while competing art historical narratives such as Marilyn Stokstad and Micheal W. Cothren's History of Art Vol
. 1 has had an estimated 10,000 copies sold since its release in 2017
.[^intro-22] Such has left Janson's History of Art as a minor player in art historical narratives toda
y.

[^intro-20]: Alexandra Peers, "Canon Fodder."

[^intro-21]: Jeffrey Weidman, "Many Are Culled but Few Are Chosen: Janson's History of Art, Its Reception, Emulators, Legacy, and Current Demise." Journal of Scholarly Publishing 38, no. 2 (January 2007): 96.
    https://muse.jhu.edu/article/209994/pdf.

[^intro-22]: "100 Best-Selling Art HIstory Books of All Time," Book Authority, https://bookauthority.org/books/best-selling-art-history-books.

+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| Edition                  | Year            | Authors                                                                                                     | Title                                                                                         | Publisher[^intro-23]                                                                   |
+==========================+=================+=============================================================================================================+===============================================================================================+========================================================================================+
| 1                        | 1963[^intro-24] | H.W. Janson with Dora Jane Janson[^intro-25]                                                                | History of Art: A Survey of the Major Visual Arts from the Dawn of History to the Present Day | Harry N. Abrams, Inc., New York                                                        |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 1 (Revised and Enlarged) | 1969            | H. W. Janson with Dora Jane Janson                                                                          | History of Art: A Survey of the Major Visual Arts from the Dawn of History to the Present Day | Prentice-Hall Inc., Englewood Cliffs, New Jersey and Harry N. Abrams, Inc., New York.  |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 2                        | 1977            | H. W. Janson with Dora Jane Janson                                                                          | History of Art: A Survey of the Major Visual Arts from the Dawn of History to the Present Day | Harry N. Abrams, Inc., New York.                                                       |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 3                        | 1986            | H. W. Janson; Revised and Expanded by Anthony F. Janson                                                     | History of Art                                                                                | Harry N. Abrams, Inc., New York and Prentice-Hall Inc., Englewood Cliffs, New Jersey.  |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 4                        | 1991            | H. W. Janson; Revised and Expanded by Anthony F. Janson                                                     | History of Art                                                                                | Harry N. Abrams, Inc., New York.                                                       |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 5                        | 1995            | H. W. Janson; Revised and Expanded by Anthony F. Janson                                                     | History of Art                                                                                | Harry N. Abrams, Inc., New York.                                                       |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 6                        | 2001            | H. W. Janson and Anthony F. Janson                                                                          | History of Art                                                                                | Harry N. Abrams, Inc., New York.                                                       |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 7                        | 2007            | Penelope J. E. Davies, Walter B. Denny, Frima Fox Hofrichter, Joseph Jacobs, Ann M. Roberts, David L. Simon | Janson's History of Art: The Western Tradition                                                | Pearson Prentice Hall, Upper Saddle River, New Jersey.                                 |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+
| 8                        | 2011            | Penelope J. E. Davies, Walter B. Denny, Frima Fox Hofrichter, Joseph Jacobs, Ann M. Roberts, David L. Simon | Janson's History of Art: The Western Tradition                                                | Pearson Prentice Hall, Upper Saddle River, New Jersey.                                 |
+--------------------------+-----------------+-------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------------+

: Edition Number, Year of Publication, Title, Authorship, and Publisher Over (as listed per Edition) Time of All Cataloged Editions of Janson's History of Art. {#tbl-janson-editions tbl-colwidths="\[8,7,30,30,25\]"}

[^intro-23]: There are discrepancies through the first and sixth editions of only Harry N. Abrams being listed as the publisher on the book itself and Harry N. Abrams and Prentice-Hall both being listed.
    Harry N. Abrams was the primary publisher through the first sixth editions as Prentice-Hall was their distributor.

[^intro-24]: I cataloged the first edition, second printing, published in 1963, but the first edition, first printing was published in 1962.
    There is no different between printings.

[^intro-25]: Dora Jane Janson is only listed upon opening the book to the title page.
    She is not listed anywhere on the exterior of the Janson's History of Art across the first edition, second printing, first edition (revised and enlarged) fourteenth printing, nor the second edition.

### Discourse of Authorship Through Editions

The primary author through the first six editions of Janson's History of Art is Horace Waldemar Janson (1913-1982), colloquially referred to as Peter.
He was born in 1913 in St. Petersburg, Russia.
His parents were of Swedish and Latvian descent.
The family moved to Hamburg, Germany after the Russian Revolution.
Janson studied at Munich and then at the University in Hamburg where he was a student of Erwin Panofsky, a prolific art historian known for his contributions to the studies of symbols and iconography in art.
In 1935, at the suggestion of his advisor, Janson fled Germany under the sponsorship of Alfred Barr, director of the Museum of Modern Art in New York.
Janson studied at Harvard between 1935 and 1942, completing a PhD in art history.
His most significant professorial position began in 1949, where he was appointed professor and later chairman of the department of art at New York University.
He remained in such a role for 25 years as he grew the department's prestige, solidifying its reputation as one of the most conservative art history programs in the country.
A decade later, in 1959, Janson issued a book titled, Key Monuments of the History of Art, in efforts to aid his undergraduates in their study of art as the availability of good personal study images were not easily accessible.
Such publication sparked Janson's History of Art released in 1962, a survey of art written with his wife, Dora Jane Janson.

\
Born Dora Jane Heineberg (1916-2002) in Philadelphia in 1916, she shared a passion for art with Peter.
The two met in Cambridge, Massachusetts as she attended Radcliffe College as an undergraduate student while Janson was studying at Harvard.
Dora Jane Janson worked with her husband on many publications, but her most notable achievement came in 1971 when she wrote an exhibition catalog for the Nasher Museum of Art at Duke on Art Nouveau jewelry, titled From Slave to Siren: The Victorian Woman and Her Jewelry from Neoclassic to Art Nouveau.
Authorship of Janson's History of Art was handed over to one of the couple's four children, Anthony F. Janson, after the sudden passing of Peter Janson in 1982.

Anthony Frederick "Tony" Janson was born March 30th, 1943 in St. Louis, Missouri.
The Jansons moved to New York in 1949, and Tony Janson frequently visited the city's museums as a child and teenager.
In 1954, the Jansons spent a year in Europe.
In an interview, Tony Janson describes his attitude upon the family's return to New York as one of " total rebellion" eventually causing his parents to search for a school that would engage him better academically.[^intro-26]
Janson details how he rediscovered a love for learning after enrolling at Riverdale Country School on a full scholarship, and it is at this time that he began to study art history.
Janson experimented as an artist himself, and cited his early affection for photography as the reason he was the first to add that medium to a survey textbook of art history.
[^intro-27] Educated at Columbia and Harvard, Janson was asked to continue the work of his father as author of Janson's History of Art
. Unable to revise the entire textbook in time, Janson "added women, which was long overdue."[^intro-28] and photography to the textbook for the third editio
n. After returning from the Vietnam War, Janson earned his PhD in art history from Harvard and went to teach at the College of Charleston before transitioning to a career as chief curator of the Indianapolis Museum of Ar
t. In 1984 Janson moved to the Ringling Museum of Art in Sarasota, Florida, and then he was Chief Curator of the North Carolina Museum of Art from 1989 to 199
3. Janson spent the rest of his career at UNC Wilmington and retired in 200
2. The last edition of Janson's History of Art that he worked on was the sixth revised, and he viewed his revisions of Janson's History of Art as more in line with his father's approach than the rest of his own scholarshi
p.

[^intro-26]: Anthony Janson, interview by Sherman Hayes, University of North Carolina Wilmington Archives and Special Collections, (March 15-19, 2007): 11.

[^intro-27]: Anthony Janson, interview by Sherman Hayes, University of North Carolina Wilmington Archives and Special Collections, (March 15-19, 2007): 7.

[^intro-28]: Anthony Janson, interview by Sherman Hayes, University of North Carolina Wilmington Archives and Special Collections, (March 15-19, 2007): 9.

The seventh and eighth editions were rewritten by a group of art historians across the United States: Penelope J. E. Davies, Walter B. Denny, Frima Fox Hofrichter, Joseph Jacobs, Ann M. Roberts, and David L. Simon.
In Table 2, each author's name, gender, ethnicity and nationality are listed, with question marks next to Frima Fox Hofrichter and Ann M. Roberts' nationality and Joseph Jacob's ethnicity as such information could not be found with full certainty.
We observe here a desire to diversify authorship in efforts by the publisher to cater to young readership, and increase sales.
Notably, there is gender diversity, with a fifty-fifty split of male to female authorship, though a complete lack of racial diversity, which is a notion reflected additionally in the text itself.
Since the Janson name has left the book's authorship, the text has declined significantly from being the most dominant art historical survey.[^intro-29]

[^intro-29]: Jeffrey Weidman, "Many are Culled but Few Are Chosen," 94.

+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Author                | Description (As listed by Publisher)                                                                                                                                                                                                                                                                             |
+=======================+==================================================================================================================================================================================================================================================================================================================+
| Penelope J. E. Davies | Associate Professor at the University of Texas, Austin. She is a scholar of Greek and Roman art and architecture as well as a field archaeologist. She is an author of Death and the Emperor: Roman Imperial Funerary Monuments from Augustus to Marcus Aurelius, and winner of the Vasari Award.                |
|                       |                                                                                                                                                                                                                                                                                                                  |
| \                     |                                                                                                                                                                                                                                                                                                                  |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Walter B. Denny       | Professor of Art History at the University of Massachusetts at Amherst. In addition to exhibition catalogues, his publications include books on Ottoman Turkish carpets, textiles, and ceramics, and articles on miniature painting, architecture and architectural decoration.                                  |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Frima Fox Hofrichter  | Professor and former Chair of the History of Art and Design department at Pratt Institute. She is the author of Judith Leyster, A Dutch Artist in Holland's Golden Age, which received CAA's Millard Meiss Publication Fund Award.                                                                               |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Joseph Jacobs         | An independent scholar, critic, and art historian of modern art in New York City. He was the curator of modern art at the John and Mable Ringling Museum of Art in Sarasota, Florida, director of the Oklahoma City Art Museum, and curator of American art at The Newark Museum, Newark, New Jersey.            |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Ann M. Roberts        | Professor of Art at Lake Forest College, she has published essays, articles and reviews on both Northern and Italian Renaissance topics. Her research focuses on women in the Renaissance, and her most recent publication is entitled Dominican Women and Renaissance Art: The Convent of San Domenico of Pisa. |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| David L. Simon        | Jetté Professor of Art at Colby College where he received the Basset Teaching Award in 2005. Among his publications is the catalogue of Spanish and southern French Romanesque sculpture in the Metropolitan Museum of Art and The Cloisters.                                                                    |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

: Authors of the Seventh and Eighth Editions of Janson's History of Art. {#tbl-janson-78 tbl-colwidths="\[25,75\]"}

### Literature Review

Many are Culled but Few are Chosen: Janson's History of Art, Its Reception, Emulators, Legacy, and Current Demise

The publication by art historian Jeffrey Weidman, released in January, 2007 in the Journal of Scholarly Publishing begins by reviewing in detail the lineage of English-language art history surveys in America, and the Jansons' text therein; discusses in detail the various reviews of the Jansons' and other survey texts, namely Helen Gardner's *Art Through the Ages*, and Marilyn Stokstad et al.'s *Art History.*[^intro-30] He concludes that Janson's History of Art functioned as the dominant art history survey through time.
Secondly, Weidman discusses the changes of Janson's History of Art between the sixth edition revised, the last edition written by Peter and Tony Janson, and the seventh edition written by the six various art historians (Table 2).
He believes that the seventh edition is a disgrace to Janson's text, and the beauty of Janson's analytically writing has been completely stripped in the seventh edition.
He also discusses, yet only through a handful of examples, artists and works which were taken out as well as added between the sixth edition revised and the seventh (for a comprehensive understanding of artists taken out and added through time refer to Fig. \_\_).
Notably, such publication does not discuss data in the aggregate when looking at who is included and excluded, and rather only looks in detail at the changes between the sixth edition revised and the seventh edition.

[^intro-30]: Jeffrey Weidman, "Many Are Culled but Few Are Chosen: Janson's History of Art, Its Reception, Emulators, Legacy, and Current Demise." Journal of Scholarly Publishing 38, no. 2 (January 2007): 85-107.
    https://muse.jhu.edu/article/209994/pdf.

Revising Art History's Big Book: Who's In and Who Comes Out?

This article published by Randy Kennedy in March of 2006 in the New York Times, also discusses the change between the sixth edition revised and the seventh edition of Janson's *History of Art*.[^intro-31]
Kennedy looks at only a few artists and works who leave and are added, particularly baffled by the exclusion of James Abbot McNeill Whistler's Arrangement in Black and Gray: The Artist's Mother.
He interviews Sarah Touburg, an editor of the seventh edition who claims that upwards of 25% of the book's content had been altered between the sixth edition revised and the seventh edition.
Interestingly, Kennedy claims that the new book adds more women and it uses art much more as a way to discuss race, class and gender.
Such an observation is not reflected in our data, which can be seen in figures \_\_, \_\_\_, and \_\_\_.
This publication also discusses the dominance and influence that Janson's History of Art has had over time, as well as its decline as the best-selling art survey in recent years.
Kennedy states that a shift in authorship was done in efforts to revitalize Janson's relevance to younger audiences.
Though this publication discusses the change between two editions, Kennedy does not do so in the aggregate.
\

[^intro-31]: Randy Kennedy, "Revising Art History's Big Book: Who's In and Who Comes Out?" New York Times, (March 7, 2006), https://www.nytimes.com/2006/03/07/arts/design/revising-art-historys-big-book- whos-in-and-who-comes-out.html.

Discussions and Depictions of Women in H.
W. Janson's History of Art, Fourth Edition

Art historian Paul E. Bolin discusses gender inequality through the first four editions of *Janson's History of Art.*[^intro-32] He states that a primary criticism of Janson's text is its lack of attention given to the importance of work of women artists.
The first two editions did not include any women, then there was an influx of female artists when Tony Janson claimed primary authorship for the publication of the third edition in 1986.
Bolin argues that even with the edition of female artists in the third and fourth editions, there is still not enough representation and that additionally the fashion in which Tony discusses female artists is apologetic and at times problematic.
Bolin uses simple statistics in his discourse, having counted a total of 9 out of the 28 new artists added to the fourth edition as female.
Bolin does not discuss racial discrimination, nor does he use complete data in the aggregate when looking gender discrimination.

[^intro-32]: Paul E. Bolin, "Discussions and Depictions of Women in H. W. Janson's History of Art, Fourth Edition," Journal of Social Theory in Art Education, Vol.
    15.(1996), pp. 146-159.
    https://core.ac.uk/download/pdf/51288476.pdf.

## Significance of Research

As far as I am aware, there has not been research done in the aggregate, quantitatively analyzing the change over time specifically regarding artist demographic in *Janson's History of Art* nor *Gardner's Art Through the Ages*.
My research is significant as it efficiently shows the gaps of representation in terms of gender, nationality, race, and ethnicity that have not been deeply touched on by previous scholars.
As mentioned in my literature review of publications that discuss *Gardner's Art Through the Ages,* the closest research I have found to my own was published by The University of Chicago Press Journals in 2020 in which two professors and library assistant out of Purdue University, randomly selected and cataloged information from the first, fifth, twelfth and sixteenth edition.
They were focused on how the length of the book changes through time, the type of image changed through time, how the percent of color images change through time, and how the number of suggested readings changes through time.
My research has more breadth, as I scraped information from all 16 editions of *Gardner's Art Through the Ages,* as well as *Janson's History of Art* and working less with how the book itself was changing, but more so how the artist's included changes through time.

Secondly, my research looks at how the area given to particular artists changes through time and which, if any, external variables can help account for the variance of such.
As mentioned in my abstract, I use a linear mixed-effects model with a random effect of the artist's name, to infer the magnitude of the space given to a particular artist divided by the area of the page of the given edition (total space ratio per page), in *Janson's History of Art* and *Gardner's Art Through the Ages,* using the potential predictor variables: Artist Race, Artist Gender, Artist Ethnicity, Artist Nationality, the count through time of the number of exhibitions an artist has at the MoMA, and the count through time of the number of exhibitions an artist has at The Whitney.
The model that best explains the variance in the total space ratio per page per artist per edition uses artist nationality and the count through time of the number of exhibitions an artist has in the MoMA as well as the interaction between those two variables.
With a log transformation applied to the outcome variable of the as it is heavily right-skewed, my linear mixed-effects model yields a conditional r squared of 53.23%.
Such denotes that 53.23% of the variance of total space ratio per page given to an artist in a given edition can be explained by the artist's nationality and the count through time of the number of exhibitions an artist has at the MoMA.
This type of statistical analysis as far as I am aware has not been done in regards to looking at how art historical survey texts evolve.
My research works to explain how the authors of *Janson's History of Art* and *Gardner's Art Through the Ages,* operate while choosing which artists are given any particular amount of space in their publications.

## Research Questions

1.  What are the demographics (artist gender, race, ethnicity, nationality) of artists included in each edition of Janson's History of Art (from 1962-2011) and Gardner's Art Through the Ages (from 1926-2020) looking at two-dimensional works after c.
    1750?

    Hypothesis: *Janson's History of Art* and *Gardner's Art Through the Ages* will include more female, nonwhite, Hispanic or Latinx artists over time; however, these artists will still be dramatically underrepresented compared to white male artists.
    Moreover, race and ethnicity will be more of a limiting factor than gender.

2.  Which variables ( artist gender, race, ethnicity, nationality, number of exhibitions at the MoMA or The Whitney) if any, predict the magnitude of an artist's inclusion in art history survey texts (*Janson's History of Art* and *Gardner's Art Through the Ages*)?

    Hypothesis: I hypothesize demographic information per artist will play a large role in accounting for the total variance in `space_ratio_per_page_total`.
    The type of race, gender, nationality, and ethnicity of an artist in theory highly impacts how much they are highlighted and discussed in introductory art history texts.
    Additionally, I hypothesize that if a given artist has many exhibitions in the MoMA or The Whitney, such would additionally account for variance in the outcome variable `space_ratio_per_page`.
    The notion behind such a hypothesis is that if an artist is being put on exhibition, they are likely to be discussed more in introductory art history texts.
